BOLD Parameter Estimation Using Particle Filters

### 2010-08-15 doc/thesis/methods.tex
mostly finished with simulation results
more work

\begin{equation}
y_{\text{gain}, 0:K} = 1.4826\text{median}_{i=0:K}(y_i - \text{median}(y_{0:K}))
\end{equation}

@@

\section{Prior Distribution}
\label{sec:PriorDist}
For the BOLD model described in \autoref{sec:BOLD Physiology}, several
different studies have endeavored to calculate parameters. The results
of these studies may be found in \autoref{tab:Params}, and the methods 
used for each may be found in \autoref{sec:Prior Works}. Unfortunately,
\cite{Friston2000} only studied regions deemed active by the General 
Linear Model; and most other research (including \cite{Friston2001}) used these results as 
the source for their priors. 
The one exception is \cite{Johnston2008}, which came to a extremely different
distributions. For a particle filter, the choice of a prior is
the single most important design choice. A very wide prior will require
more particles to be sufficiently dense, a very thin (low variance) prior may miss
the true parameters. Consequently, for this work it was natural
to use priors that will give results consistent with previous works, 
\cite{Friston2000}. This constrains the usefulness of the model to
areas that fall within the prior distribution, yet will allow results
to be comparable to other works. There is a significant need for better
estimates of the physiological parameters; and, while physical experiments
may not be possible, it would not be unreasonable to do a study with
exhaustive simulated annealing or hill climbing tests for multiple
regions and multiple patients.

There is an interesting anomaly with the priors found in virtually all
the works that characterized the parameters, except \cite{Johnston2008}.
The BOLD signal is universally recognized to be around $2-3\%$, maybe
reaching $5\%$ in extreme activation. Yet using the mean priors
from \cite{Friston2000}, the signal response for a $.1$ second
impulse only reaches maybe half a percent, as \autoref{fig:MeanResponseF}
shows.

\begin{figure}
\centering
\label{fig:MeanResponseF}
\includegraphics[trim=6cm 3cm 6cm 3cm,width=16cm]{images/mean_response}
\caption{Response to $0.1s$ impulses with the mean parameters from \cite{Friston2000}}
\end{figure}

@@

\begin{figure}
\centering
\label{fig:MeanResponseJ}
\includegraphics[trim=6cm 3cm 6cm 3cm,width=16cm]{images/mean_response_johnston}
\caption{Response to $0.1s$ impulses with the mean parameters from \cite{Johnston2008}}
\end{figure}

@@

\section{Resampling}
\label{sec:Resampling}
The algorithm for resampling is described in \autoref{sec:Particle Filter Resampling}.
When regularizing, the Gaussian kernel is convenient,
because it is simple to sample from and long tailed.
As discussed in \autoref{sec:Particle Filter Resampling},
as long as resampling is kept as a last resort, some over-smoothing
doesn't impair convergence. Therefore, for this work I chose a Gaussian kernel of
bandwidth equal to the original distribution's covariance. Obviously this will
apply a rather large amount of smoothing to the distribution; however, on average
resampling is only applied every 20 to 30 measurements, and because randomization
is being applied to model updates this gives the filter some mobility.

@@

\section{Choosing $P(y_k | x_k)$}
\label{sec:Methods Weighting Function}
The choice of $P(y_k | x_k)$ is the second most important design decision, behind 
the prior. While the conventional choice for an unknown distribution is the 
Gaussian, there are reasons why it may not be the best in this case.  
As noted in \autoref{sec:Introduction Noise}, the noise is not strictly Gaussian,
nor is it strictly Wiener. As with any unknown noise however, it is necessary 
to make some assumption. If the weighting function ($P(y_k | x_k)$) exactly
matches the measurement error, then the ideal particle filter will result.
Particles with $x_k$'s that repeatedly estimate $y_k$ with large residual 
will quickly have weights near 0. Thus, a weighting function that
exactly matches $P(Y(t) | X(t))$ will easily, and correctly throw out incorrect
particles.  The cost of choosing an overly broad distribution for this
function is slow convergence.  On the other hand, an overly thin 
distribution will lead to particle deprivation (all particles
being zero-weighted).  I tested several weighting
functions: in addition to the Gaussian I also tested the Laplace and Cauchy
distributions, both of which have much wider tales than the Gaussian. 
Wider tailed distributions don't down-weight
particles as fast; and converge more slowly (and perhaps more accurately). 
The Laplace distribution also has the
benefit of a non-zero slope at the origin; which means that even
it will distinguish between particles even near the origin.

>>>>

\begin{equation}
y_{\text{gain}, 0:K} = 1.4826\text{median}_{i=0:K}(y_i - \text{median}(y_{0:K}))
\label{eq:mad}
\end{equation}

@@

\subsection{Prior Distribution}
\label{sec:PriorDist}
For the BOLD model described in \autoref{sec:BOLD Physiology}, several
different studies have endeavored to calculate parameters. The results
of these studies may be found in \autoref{tab:Params}, and the methods 
used for each may be found in \autoref{sec:Prior Works}. Unfortunately,
\cite{Friston2000} only studied regions deemed active by the General 
Linear Model; and most other research (including \cite{Friston2001}) used these results as 
the source for their priors. 
The one exception is \cite{Johnston2008}, which came to a extremely different
distributions. For a particle filter, the choice of a prior is
the single most important design choice. A very wide prior will require
more particles to be sufficiently dense, a very thin (low variance) prior may miss
the true parameters. Consequently, for this work it was natural
to use priors that will give results consistent with previous works, 
\cite{Friston2000}. This constrains the usefulness of the model to
areas that fall within the prior distribution, yet will allow results
to be comparable to other works. There is a significant need for better
estimates of the physiological parameters; and, while physical experiments
may not be possible, it would not be unreasonable to do a study with
exhaustive simulated annealing or hill climbing tests for multiple
regions and multiple patients.

There is an interesting anomaly with the priors found in virtually all
the works that characterized the parameters, except \cite{Johnston2008}.
The BOLD signal is universally recognized to be around $2-3\%$, maybe
reaching $5\%$ in extreme activation. Yet using the mean priors
from \cite{Friston2000}, the signal response for a $0.1$ second
impulse only reaches half a percent, as \autoref{fig:MeanResponseF}
shows.

\begin{figure}
\centering
\includegraphics[trim=6cm 3cm 6cm 3cm,width=16cm]{images/mean_response}
\caption{Response to $0.1s$ impulses with the mean parameters from \cite{Friston2000}}
\label{fig:MeanResponseF}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[trim=6cm 3cm 6cm 3cm,width=16cm]{images/mean_response_johnston}
\caption{Response to $0.1s$ impulses with the mean parameters from \cite{Johnston2008}}
\label{fig:MeanResponseJ}
\end{figure}

@@

\subsection{Resampling}
\label{sec:Resampling}
The algorithm for resampling is described in \autoref{sec:Particle Filter Resampling}.
When regularizing, the Gaussian kernel is convenient,
because it is simple to sample from and long tailed.
As discussed in \autoref{sec:Particle Filter Resampling},
as long as resampling is kept as a last resort, some over-smoothing
doesn't impair convergence. Therefore, for this work I chose a Gaussian kernel of
bandwidth equal to the original distribution's covariance. Obviously this will
apply a rather large amount of smoothing to the distribution; however, on average
resampling is only applied every 20 to 30 measurements, and because randomization
is being applied to model updates this gives the filter some mobility.

@@

\subsection{Choosing $P(y_k | x_k)$}
\label{sec:Methods Weighting Function}
The choice of $P(y_k | x_k)$ is the second most important design decision, behind 
the prior. While the conventional choice for an unknown distribution is the 
Gaussian, there are reasons why it may not be the best in this case.  
As noted in \autoref{sec:Introduction Noise}, the noise is not strictly Gaussian,
nor is it strictly Wiener. As with any unknown noise however, it is necessary 
to make some assumption. If the weighting function ($P(y_k | x_k)$) exactly
matches the measurement error, then the ideal particle filter will result.
Particles with $x_k$'s that repeatedly estimate $y_k$ with large residual 
will quickly have weights near 0. Thus, a weighting function that
exactly matches $P(Y(t) | X(t))$ will easily, and correctly throw out incorrect
particles.  The cost of choosing an overly broad distribution for this
function is slow convergence.  On the other hand, an overly thin 
distribution will lead to particle deprivation (all particles
being zero-weighted).  I tested several weighting
functions: in addition to the Gaussian I also tested the Laplace and Cauchy
distributions, both of which have much wider tales than the Gaussian. 
Wider tailed distributions don't down-weight
particles as fast; and converge more slowly (and perhaps more accurately). 
The Laplace distribution also has the
benefit of a non-zero slope at the origin; which means that even
it will distinguish between particles even near the origin.

### 2010-08-15 doc/thesis/results.tex
more work

Note that except for \autoref{sec:VeryLongSim} the same stimulus sequence
that was used in real data (\autoref{sec:RealData}) was used in the simulations. This sequence may
be found in \autoref{tab:flashes}, but because of the availability of ground
truth in simulations is not displayed in this section.

>>>>

Note that except for \autoref{sec:VeryLongSim} the same stimulus sequence
that was used in real data (\autoref{sec:RealData}) was used in the simulations. This sequence may
be found in \autoref{sec:ExperimentConfig}, but because of the availability of ground
truth in simulations is not displayed in this section.

### 2010-08-15 doc/thesis/resultsB.tex
more work

%scanner info?
\section{FMRI Configuration}
For the FMRI data discussed in \autoref{sec:RealData}, tests were 
performed on a right handed volunteer using a GE SIGNA HDx 1.5 Tesla
scanner with a single echo EPI sequence. Slice spacing was
5mm, and pixel sizes were 3.75mm.  Repitition Time was 2.1s, 
Echo Time was 40 ms and the imaging frequency was set to $63.854$MHz.
The image resolution was $64 \times 64 \times 28$. 
The subject was presented with either a single or
double flash and was asked to respond with a right handed
or lefted handed finger tap, respectively. The FMRI began 18.9 seconds
before the beginning of the experiment, to allow for transients 
in to image to settle out. The timing of the flashes are shown 
below with time 0 corresponding to the beginning of
the 10th TR (so 9 images were dropped).

>>>>

%scanner info?
\section{Experiment Configuration}
\label{sec:ExperimentConfig}
For the FMRI data discussed in \autoref{sec:RealData}, tests were 
performed on a right handed volunteer using a GE SIGNA HDx 1.5 Tesla
scanner with a single echo EPI sequence. Slice spacing was
5mm, and pixel sizes were 3.75mm.  Repitition Time was 2.1s, 
Echo Time was 40 ms and the imaging frequency was set to $63.854$MHz.
The image resolution was $64 \times 64 \times 28$. 
The subject was presented with either a single or
double flash and was asked to respond with a right handed
or lefted handed finger tap, respectively. The FMRI began 18.9 seconds
before the beginning of the experiment, to allow for transients 
in to image to settle out. The timing of the flashes are shown 
below with time 0 corresponding to the beginning of
the 10th TR (so 9 images were dropped).
